Handling String-Number Conversion in String Constraint Solving

### 2019-11-09 draft.tex
polishing 5.1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametric Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
\todo{add Presburger constraints to allow more expressive power.}
\lh{TODO: solve the where to add $-1$ problem}
%We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. 
%Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
%We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
%An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

The crucial feature of parametric flat automata is that their semantics can be faithfully represented by a linear integer arithmetic formulae and handled by an SMT solver, allowing to circumvent more costly automata algorithms and decision procedures. \lh{concretize and justify this better?}
%
Achieving this was the purpose of introducing the structural constraints of PFA. 
Their implication central for our method is that every parametric string $x\in L(A)$ is uniquelly determined by its Parikh image $\parikhof{x}$, as formalised by the following lemma.

@@

\paragraph{Synchronisation of PFA}
The following construction of the so called synchronisation formula of two PFA $A$ and $A'$ over character variables $V$ and $V'$, respectively, is a crucial component of building underapproximations of string constraints. 
Synchronisation formula is a formula $\syncfof A {A'}$ over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. Its representation of the intersection includes an additional information about the synchronisation of $A$ and ${A'}$, also needed for building underapproximations of string constraints. 
Namely, it characterises the set of pairs of string encoding encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$.% which constitute the intersection.

Let $A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
We will build their asynchronous product, an automaton $\syncof A {A'}$, 
and then extract the synchronisation formula $\syncfof A {A'}$ from it.
%
The product automaton uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ will correspond to a pair of accepting runs, 
a run of $A$ over a parametric string $x$ and a run of ${A'}$ over a parametric string $x'$. 
The word accepted by the run of $\syncof A {A'}$ will induce the constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric strings have the same interpretation, i.e. $I(x) = I(x')$.

@@

Formally, the product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
%The we define of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
The core part of the synchronisation formula is the Parikh formula $\parikhfof {\syncof A {A'}}$ of the product, which encodes all its runs.
%
To extract form a run of $\syncof A {A'}$ the corresponding run of $A$ and of $A'$, we use the following constraint:
$$ \Psi_{\#} \defeq 
\left(\bigwedge_{v\in\cvarone}\#v = \sum_{x' \in \cvartwo_\epsilon} \#{(v,x')}\right)
\land
\left(\bigwedge_{v'\in\cvartwo} \#v' = \sum_{x \in \cvarone_\epsilon} \#{(x,v')}\right)
$$
Notice that $x,x'$ are either variables or $\epsilon$.
%\begin{enumerate}
%	\item for all $v \in \cvarone$, $\#(v) = \sum_{v' \in \cvartwo} \#{(v,v')}$   
%	\item for all $v \in \cvartwo$, $\#(v) = \sum_{v' \in \cvarone} \#{(v',v)}$ 
%\end{enumerate}
Finally, to ensure that the interpretations of the parametric strings accepted by $A$ and $A'$ are the same, the following constraint is used:
% condition 1. and 2. required for the pair of symbolic string $(w_n, w_m)$ and $I$, i.e., $I(w_n) = I(w_m)$ and $I$ assign the same value to all versions of the same variable (we use $-1$ as the value of $\epsilon$). 
$$ \Psi_= \defeq
\bigwedge_{x\in\cvarone_\epsilon,x'\in\cvartwo_\epsilon} \#(x,x')>0 \rightarrow (x=x')
$$
%\begin{enumerate}
%	\item for all $v\in \cvarone$ and $v'\in \cvartwo$, $\#{(v,v')}>0 \rightarrow (v=v')$, 
%	\item for all $v\in \cvarone$, $\#{(v,\epsilon)}>0 \rightarrow (v=\epsilon)$, and
%	\item for all $v\in \cvartwo$,$\#{(\epsilon,v')}>0 \rightarrow (v'=\epsilon)$.
%\end{enumerate}
The synchronisation formula is the conjunction 
$$\label{lemma:synccorrect}
%\syncfof A {A'} = \exists \pvarsof{\cvarone\times\cvartwo}\exists {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
\syncfof A {A'} \defeq \parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$
The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect}. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\exists V_\times . \syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Informally, the models of $\syncfof A {A'}$ encode precisely the pairs of equivalent encodings of words from $\semof{A}$ and $\semof{B}$, that constitute the intersection $\semof{A} \cap \semof{B}$. A minor technical difficulty is that the modes of $\syncfof A {A'}$ include also an assignment to the auxiliary variables of $V_\times = (\cvarone_\epsilon\times\cvartwo_\epsilon) \cup \pvarsof{(\cvarone_\epsilon\times\cvartwo_\epsilon)}$, hence in the lemma we have to restrict the models only to the original variables $\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo$, which we achieve by quantifying out $V_\times$.

@@

\lh{TODO: put this somewhere and explain}

\begin{lemma}
	Symbolic flat automata are closed under concatenation.
\end{lemma}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametric Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
\todo{additional presburget constraints to encode $\epsilon$ added at translation to a formula}
%We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. 
%Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
%We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
%An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

The crucial feature of parametric flat automata is that their semantics can be faithfully represented by a linear integer arithmetic formulae and handled efficiently by an SMT solver. 
\lh{, allowing to circumvent more costly automata algorithms and decision procedures. Concretize and justify this better?}
%
Achieving this was the purpose of introducing the structural constraints of PFA. 
Their implication central for our method is that every parametric string $x\in L(A)$ is uniquelly determined by its Parikh image $\parikhof{x}$, as formalised by the following lemma.

@@

\paragraph{Synchronisation of PFA}
The following construction is an important step in building underapproximations of string constraints. 
For two PFA $A$ and $A'$ over character variables $V$ and $V'$, respectively, 
it builds \emph{synchronisation formula} $\syncfof A {A'}$, a linear formula over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. 
Besides the intersection itself, it also represents the set of all pairs of string encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$ which constitute the intersection.

Particularly, a synchronisation formula is build for two PFA  
$A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
%
It is extracted from the asynchronous product of $A$ and $A'$, an automaton $\syncof A {A'}$. 
%
The asynchronour product is an automaton that uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to a pair of accepting runs, 
a run of $A$ over a parametric string $x$ and a run of ${A'}$ over a parametric string $x'$. 
The word accepted by the run of $\syncof A {A'}$ induces constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric strings have the same interpretation, i.e. $I(x) = I(x')$.

@@

Formally, the asynchronous product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
%The we define of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
The synchronisation formula $\syncof A {A'}$ is extracted from $\syncof A {A'}$ as follows.
Its first, core part is the Parikh formula $\parikhfof {\syncof A {A'}}$ of the product, which encodes all runs of the product.
%
The second part is a constraint that extracts from a run of $\syncof A {A'}$ the corresponding runs of $A$ and of $A'$:
$$ \Psi_{\#} \defeq 
\left(\bigwedge_{v\in\cvarone}\#v = \sum_{x' \in \cvartwo_\epsilon} \#{(v,x')}\right)
\land
\left(\bigwedge_{v'\in\cvartwo} \#v' = \sum_{x \in \cvarone_\epsilon} \#{(x,v')}\right)
$$
Notice that $x,x'$ are either variables or $\epsilon$.
%\begin{enumerate}
%	\item for all $v \in \cvarone$, $\#(v) = \sum_{v' \in \cvartwo} \#{(v,v')}$   
%	\item for all $v \in \cvartwo$, $\#(v) = \sum_{v' \in \cvarone} \#{(v',v)}$ 
%\end{enumerate}
Finally, the third part forces the interpretations of the parametric strings accepted by $A$ and $A'$ be the same:
% condition 1. and 2. required for the pair of symbolic string $(w_n, w_m)$ and $I$, i.e., $I(w_n) = I(w_m)$ and $I$ assign the same value to all versions of the same variable (we use $-1$ as the value of $\epsilon$). 
$$ \Psi_= \defeq
\bigwedge_{x\in\cvarone_\epsilon,x'\in\cvartwo_\epsilon} \#(x,x')>0 \rightarrow (x=x')
$$
%\begin{enumerate}
%	\item for all $v\in \cvarone$ and $v'\in \cvartwo$, $\#{(v,v')}>0 \rightarrow (v=v')$, 
%	\item for all $v\in \cvarone$, $\#{(v,\epsilon)}>0 \rightarrow (v=\epsilon)$, and
%	\item for all $v\in \cvartwo$,$\#{(\epsilon,v')}>0 \rightarrow (v'=\epsilon)$.
%\end{enumerate}
The synchronisation formula is the conjunction 
$$
%\syncfof A {A'} = \exists \pvarsof{\cvarone\times\cvartwo}\exists {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
\syncfof A {A'} \defeq \parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$
The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect}. The Lemma can also be understood as a precise specification the semantics of the synchronisation formula. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\exists V_\times . \syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Informally, the lemma states that the models of $\syncfof A {A'}$ encode precisely the pairs of equivalent encodings of words from $\semof{A}$ and $\semof{B}$, that constitute the intersection $\semof{A} \cap \semof{B}$. A minor technical difficulty is that the modes of $\syncfof A {A'}$ include also an assignment to the auxiliary variables of $V_\times = (\cvarone_\epsilon\times\cvartwo_\epsilon) \cup \pvarsof{(\cvarone_\epsilon\times\cvartwo_\epsilon)}$. Hence in the lemma, we have to restrict the models to the original variables $\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo$. This is  achieve by quantifying out the variables of $V_\times$.

@@

\lh{removed the closed under concatenation lemma, no more true due to epsilons}

%\begin{lemma}
%	Symbolic flat automata are closed under concatenation.
%\end{lemma}
